A Large-scale Study of Polymorphism in R

### 2019-03-22 notebooks/corpus_stats_and_plots.Rmd
results, some tables and stuff

plot_both <- ggplot( data=zzz, aes(package_name, value, group=1))
plot_both <- plot_both + geom_line(aes(colour = language), size=0.001) # +  geom_point(aes(colour = language))
plot_both <- plot_both + labs(
  title = "Analysis of R Corpus (R and C)",
  y = "Lines of Code in Package (Log10-scaled)"
)

plot_both <- plot_both + theme(axis.text.x = element_blank(), axis.title.x=element_blank(), axis.ticks.x=element_blank()) + scale_y_continuous(name="Lines of Code in Package") # , breaks = c(-4, -3, -2, -1, 0, 1, 2, 3, 4), labels = function(x) abs(x^10)

@@

ggsave("../paper/plots/linesofrandccode.pdf", plot_both)

@@

records_by_fun <- unlist(readRDS("../data/total_list_of_records_by_pkg.RDS"))
records_by_fun_df <- data.frame(pname=names(records_by_fun), records=unname(records_by_fun))
# records_by_fun_df <- records_by_fun_df[order(records_by_fun_df$records), ]
# records_by_fun_df$pname <- factor(records_by_fun_df$pname, levels=records_by_fun_df$pname)

@@

# generate file
ggsave("../paper/plots/recordsbypkg.pdf", plot_records)
```

>>>>

plot_both <- ggplot( data=zzz, aes(package_name, value, group=1))
plot_both <- plot_both + geom_line(aes(colour = language), size=0.05) # +  geom_point(aes(colour = language))
plot_both <- plot_both + labs(
  title = "Analysis of R Corpus (R and C)",
  y = "Lines of Code in Package (Log10-scaled)"
) + guides(colour=F)

# cut_top <- data.frame( x = c(-Inf, Inf), y = 2, cutoff = factor(2) )
# cut_bot <- data.frame( x = c(-Inf, Inf), y = -2, cutoff = factor(-2) )

plot_both <- plot_both + theme(
  axis.text.x  = element_blank(), 
  axis.title.x = element_blank(),
  axis.ticks.x = element_blank(),
  axis.ticks.y = element_line(), 
  panel.grid.major.y = element_line(colour = "red")) + # ,
  # panel.ontop = T) + 
  scale_y_continuous(name="Lines of Code in Package", breaks = c(-4, -3, -2, -1, 0, 1, 2, 3, 4), 
                     labels = c(10000, 1000, 100, 10, 0, 10, 100, 1000, 10000))
  # geom_line(aes( x, y), cut_top) +
  # geom_line(aes( x, y), cut_bot)

@@

ggsave("../paper/plots/linesofrandccode.jpg", plot_both)

@@

records_by_fun <- unlist(readRDS("../data/total_list_of_records_by_pkg.RDS"))
records_by_fun_df <- data.frame(pname=names(records_by_fun), records=unname(records_by_fun)/1000)
# records_by_fun_df <- records_by_fun_df[order(records_by_fun_df$records), ]
# records_by_fun_df$pname <- factor(records_by_fun_df$pname, levels=records_by_fun_df$pname)

@@

# generate file
ggsave("../paper/plots/recordsbypkg.jpg", plot_records)
```

### 2019-03-22 paper/main.tex
results, some tables and stuff

\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=0.6]{linesofrandccode.pdf}
\caption{Lines of Code (R and C) in the Analyzed Corpus}
\label{allcloc}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=0.6]{recordsbypkg}
\caption{Numbers of Recorded Function Invocations in the Analyzed Corpus}
\label{allcloc}
\end{center}
\end{figure}

@@

In this section, we will detail our methodology for collecting data.  The
basic idea we are exploring is to observe arguments and return values of
function calls, and from these generalize possible type signatures for the
called functions. To this end, we build on an open source tool called
\genthat whose purpose was to generate unit tests for R
libraries~\cite{issta18}.  To generate test cases, one must record and
recreate accurately arguments and return values and avoid generating
redundant tests (i.e. tests that do not increase code coverage). For our
purposes, we change the existing tool in two main ways, we record argument
\emph{shapes} rather than their values, given the limited number of shapes
this allowed us to avoid having to run code coverage minimization phase of
the tool. Both changes were beneficial for scalability -- allowing us to
trace more calls than was reported in~\cite{issta18}.

@@

To illustrate our approach consider the following script which adds a double
to an integer, and then creates a matrix and a vector, adds them together to
yield a new matrix.

@@

The functions being called here are \k{`+`}, \k{`<-`}, \k{c} and
\k{`attr<-`}.  Some functions syntactic support, addition is infix, and the
assignment to an attribute is written mixfix. The shapes we would expect to
record are as follows:

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
`+`: D[1], I[1] -> D[1]
c: D[1], D[1], D[1], D[1] -> D[4]
`<-`: S, D[3] -> D[3]
c: D[1], D[1] -> D[2]
`attr<-1`:  S, C, D[2] -> 
c: D[1], D[1] -> D[2]
`+`: D[4]{dim=D[2]}, D[2] -> D[4]{dim=D[2]}
\end{lstlisting}}\caption{Recorded shapes}\label{shapes}\end{figure}

@@

We will need a notion of polymorphism when undertaking this analysis.  As R
is unityped, and the only notion of polymorphism in the language is its use
of dynamic dispatch on the class of an argument.  For our purposes, we will
define 3 simple types of polymorphism, closely connected with the reflection
functions we use.  We start by defining what it means for an {\it argument}
to be polymorphic.  \AT{Could be valuable to have an example here.}

@@

\item attributes are a bit more tricky.  We can't simply count the number of
  attributes that have inhabited an argument, since values can contain
  arbitrarily many values (and having several attributes isn't
  polymorphism---think records).  Instead, we define the {\it attribute
    pattern} of a value to be the set of all attribute name and type pairs
  in the attribute list of a value (obtained via \code{attributes}).  And so
  an argument is {\it polymorphic in attribute} if it has been called with
  at least two distinct attribute patterns.

@@

\item to ease dealing with large sorted sets of non-numeric values, R offers
  the \code{factor} and \code{levels} functions.  \code{factor(x,
    levels=someOrder)} adds a {\tt "levels"} attribute to \code{x}, which
  specifies an ordering for \code{x} according to the \code{sortedOrder}
  parameter.  \code{factor} also changes the class of the parameter being
  factored to \code{"factor"}.

@@

The logical ``baseline'' for any design of a static type system for R is
promotion of the language's dynamic types (in the sense of \code{typeof}) to
static types, but there are some clear limitations to this approach.  For
example, \code{typeof} makes no distinction between vectors and scalars, as
indeed there is no such distinction in R.  But that doesn't mean that there
{\it shouldn't} be a distinction, as an all-new runtime environment design
for R might benefit from such a distinction.

To address these limitations, we augment the logical naive type system with
more information which was alluded to in the discussion of our data
collection methodology.  We will consider unit-length vectors to be scalars,
collect list member type information to build an accurate list type, and
promote NAs to their own type.  In addition, we collect attribute and class
information to distinguish a type for data frames and matrices, as the
\code{typeof} these is simply {\tt list} in R.  This type system is referred
to as the Fine-Grained Type System, or FGTS.  An example of a signature in
this type system can be found in Figure~\AT{figggg}.

The first question we will address is: {\it how polymorphic is R?}  To
answer that, we will first look at the number of monomorphic arguments
across all of the packages we analyzed.  The data can be found in
Figure~\AT{fiGgGgGgGGg}.

@@

We see that roughly 10\% of all argument positions in R are polymorphic.
The numbers for attribute and class polymorphism are included for
completeness, but the important number is that of type polymorphism (as our
method of constructing types sometimes factors in class and attribute).
Specifically for classes, often the class and \code{typeof} a value will
line up (refer to Section~\ref{sec:rlang}).

As for functions, roughly 25\% are polymorphic, or put differently 25\% have at least one polymorphic argument or a polymorphic return.
The discrepancy between this and the polymorphicity of arguments is explained by the tendency of some functions, particularly those dealing with models, to have {\it many} arguments.

>>>>

\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=0.15]{linesofrandccode.jpg}
\caption{Lines of code in the analyzed corpus. \# of lines of R code are above 0, \# of lines of C code are below 0.}
\label{allcloc}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=0.15]{recordsbypkg}
\caption{Numbers of Recorded Function Invocations in the Analyzed Corpus}
\label{allcloc}
\end{center}
\end{figure}

@@

In this section, we will detail our methodology for collecting data.  The
basic idea we are exploring is to observe arguments and return values of
function calls, and from these generalize possible type signatures for the
called functions. To this end, we build on an open source tool called
\genthat whose purpose is to generate unit tests for R
libraries~\cite{issta18}.  To generate test cases, one must record and
accurately recreate arguments and return values and avoid generating
redundant tests (i.e. tests that do not increase code coverage). For our
purposes, we change the existing tool in two main ways: We record argument
\emph{shapes} rather than their values, and given the limited number of shapes
this allowed us to avoid having to run the code coverage minimization phase of
the tool. Both changes were beneficial for scalability---allowing us to
trace more calls than was reported in~\cite{issta18}.

@@

To illustrate our approach, consider the following script which adds a double
to an integer, and then creates a matrix and a vector, and adds them together to
get a new matrix.

@@

The functions being called here are \k{`+`}, \k{`<-`}, \k{c} and
\k{`attr<-`}.  Some functions syntactic support \AT{?}, addition is infix, and the
assignment to an attribute is written mixfix. The shapes we would expect to
record are as follows:

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
`+`: D[1], I[1] -> D[1] # (args: double size 1, integer size 1, return: double size 1)
c: D[1], D[1], D[1], D[1] -> D[4]
`<-`: S, D[3] -> D[3]
c: D[1], D[1] -> D[2]
`attr<-1`:  S, C, D[2] -> 
c: D[1], D[1] -> D[2]
`+`: D[4]{dim=D[2]}, D[2] -> D[4]{dim=D[2]}
\end{lstlisting}}\caption{Recorded shapes}\label{shapes}\end{figure}

@@

We will need a notion of polymorphism when undertaking this analysis. R
is unityped, and the only notion of polymorphism in the language is its use
of dynamic dispatch on the class of an argument.  For our purposes, we will
define 3 simple types of polymorphism, closely connected with the reflection
functions we use.  We start by defining what it means for an {\it argument}
to be polymorphic.  \AT{Could be valuable to have an example here.}

@@

\item attributes are a bit more tricky.  We can't simply count the number of
  attributes that have inhabited an argument, since values can contain
  arbitrarily many attributes (and having several attributes isn't
  polymorphism---think records).  Instead, we define the {\it attribute
    pattern} of a value to be the set of all attribute name and type pairs
  in the attribute list of a value (obtained via \code{attributes}).  And so
  an argument is {\it polymorphic in attribute} if it has been called with
  at least two distinct attribute patterns.

@@

\item to ease dealing with large sorted sets of non-numeric values, R offers
  the \code{factor} and \code{levels} functions.  \code{factor(x,
    levels=someOrder)} adds a {\tt "levels"} attribute to \code{x}, which
  specifies an ordering for \code{x} according to the \code{sortedOrder}
  parameter.  \code{factor} also changes the class of the parameter being
  factored to \texttt{"factor"}.

@@

The logical ``baseline'' for the design of a static type system for R is
promotion of the language's dynamic types (in the sense of \code{typeof}) to
static types, but there are some clear limitations to this approach.  For
example, \code{typeof} makes no distinction between vectors and scalars, as
indeed there is no such distinction in R.  But that doesn't mean that there
{\it couldn't} be a distinction, as an all-new runtime environment design
for R might benefit from e.g. not needing to vectorize scalar values.

To address these limitations, we augment the logical naive type system with
more information which was alluded to in the discussion of our data
collection methodology.  We will consider unit-length vectors to be scalars,
collect list member type information to build an accurate list type, and
promote NAs to their own type.  In addition, we collect attribute and class
information to distinguish a type for data frames and matrices, as the
\code{typeof} these is simply {\tt list} in R.  This type system is referred
to as the Fine-Grained Type System, or FGTS.  An example of a signature in
this type system can be found in Figure~\ref{fig:exFGTS}.

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
> cat_and_list <- function(x, y) as.list(c(x, y)) # c(2, c(3, 1)) => c(2, 3, 1)
> cat_and_list(2, c(3, 1)) # => list(2, 3, 1)
\end{lstlisting}}
\begin{tabular}{@{}r|l|l|l@{}}\hline
\bf Argument & \bf Type & \bf Class &\bf Attributes \\
x & \tt scalar/double & \tt numeric & \tt \{\} \\
y & \tt vector/double & \tt numeric & \tt \{\} \\
retv & \tt list<double> & \tt list & \tt \{\}  
\end{tabular}
\caption{Example call and FGTS signature}\label{fig:exFGTS}\end{figure}

The first question we will address is: {\it how polymorphic is R?}  To
answer that, we will first look at the number of monomorphic arguments
across all of the packages we analyzed.  The data can be found in
Tables~\ref{tab:argcountsFGTS}~and~\ref{tab:funcountsFGTS}.
\AT{TODO: have these tables be generated automatically}
\AT{TODO: side by side in minipage?}

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 18:03:30 2019
\begin{table}[ht]
\label{tab:argcountsFGTS}
\centering
\begin{tabular}{rlrr}
  \hline
 & Argument Polymorphism & Count & Percentage \\ 
  \hline
1 & Full Monomorphic & 448K & 85.50 \\ 
  2 & Monomorphic in Type & 471K & 90.28 \\ 
  3 & Monomorphic in Class & 480K & 92.16 \\ 
  4 & Monomorphic in Attribute Pattern & 480K & 91.99 \\ 
  5 & Total Seen & 521K & -- \\ 
   \hline
\end{tabular}
\caption{Account of {\it argument} polymorphism in the FGTS.}
\end{table}

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 18:10:46 2019
\begin{table}[ht]
\label{tab:funcountsFGTS}
\centering
\begin{tabular}{rlrr}
  \hline
 & Function Polymorphism & Count & Percentage \\ 
  \hline
1 & Full Monomorphic & 82K & 66.27 \\ 
  2 & Monomorphic in Type & 94K & 75.74 \\ 
  3 & Monomorphic in Class & 97K & 78.45 \\ 
  4 & Monomorphic in Attribute Pattern & 96K & 77.47 \\ 
  5 & Total Seen & 124K & -- \\ 
   \hline
\end{tabular}
\caption{Account of {\it function} polymorphism in the FGTS.}
\end{table}

@@

In the tables, ``Full Monomorphic'' denotes monomorphism in type, class, and attribute.
We see that roughly 10\% of all argument positions in R are polymorphic in type.
The numbers for attribute and class polymorphism are included for
completeness, but the important number is that of type polymorphism (as our
method of constructing types sometimes factors in class and attribute).
Specifically for classes, often the class and \code{typeof} a value will
line up (refer to Section~\ref{sec:rlang}).

As for functions, roughly 25\% are polymorphic in type, or put differently 25\% have at least one polymorphic argument or a polymorphic return.
The discrepancy between this and the polymorphicity of arguments is explained by the tendency of some functions, particularly those dealing with models, to have {\it many} arguments.
